---
title: "Getting Started with regexpert"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with regexpert}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(regexpert)
library(magrittr)
```

## The Problem with Regular Expressions

If you've ever needed to extract a phone number or pull out specific text in R, you've probably run into something like this:
```{r eval = FALSE}
stringr::str_extract_all("Call us at 800-555-1234", "\\d+")
```

That `"\\d+"` is a regular expression. It works — but what does it *mean*? For anyone who doesn't have regex memorized, this is a wall. You have to stop what you're doing, look up the syntax, and hope you get the escaping right.

This is the problem `regexpert` solves.

## The Build → Act Workflow

`regexpert` breaks pattern matching into two plain-English steps:

1. **Build** your pattern using a descriptive `xp_build_*` function.
2. **Act** on it using `xp_find()` to return matches.

These steps are connected with the pipe (`%>%` or `|>`), so your code reads like a sentence:
```{r eval = FALSE}
"Call us at 800-555-1234" %>%
  xp_build_digits() %>%
  xp_find()
#> [1] "800" "555" "1234"
```

No regex required. You described what you wanted in plain English, and `regexpert` handled the rest.

## The Builders

### `xp_build_digits()`

Matches any numeric character (0–9).
```{r eval = FALSE}
"Order #88412 placed on 2024-01-15" %>%
  xp_build_digits() %>%
  xp_find()
#> [1] "88412" "2024" "01" "15"
```

Use `negate = TRUE` to find everything *except* digits:
```{r eval = FALSE}
"Price: $100!" %>%
  xp_build_digits(negate = TRUE) %>%
  xp_find()
#> [1] "Price: $" "!"
```

### `xp_build_letters()`

Matches any alphabetic character (a–z, A–Z).
```{r eval = FALSE}
"Order #88412 placed" %>%
  xp_build_letters() %>%
  xp_find()
#> [1] "Order" "placed"
```

Use `negate = TRUE` to find everything *except* letters:
```{r eval = FALSE}
"abc123" %>%
  xp_build_letters(negate = TRUE) %>%
  xp_find()
#> [1] "123"
```

### `xp_build_whitespace()`

Matches spaces, tabs, and newlines.
```{r eval = FALSE}
"hello   world" %>%
  xp_build_whitespace() %>%
  xp_find()
#> [1] " " " " " "
```

Use `negate = TRUE` to find everything *except* whitespace:
```{r eval = FALSE}
"hello   world" %>%
  xp_build_whitespace(negate = TRUE) %>%
  xp_find()
#> [1] "hello" "world"
```