---
title: "Getting Started with regexpert"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with regexpert}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(regexpert)
library(magrittr)
```

## The Problem with Regular Expressions

If you've ever needed to extract a phone number or pull out specific text in R, you've probably run into something like this:
```{r eval = FALSE}
stringr::str_extract_all("Call us at 800-555-1234", "\\d+")
```

That `"\\d+"` is a regular expression. It works — but what does it *mean*? For anyone who doesn't have regex memorized, this is a wall. You have to stop what you're doing, look up the syntax, and hope you get the escaping right.

This is the problem `regexpert` solves.

## The Build → Operate → Act Workflow

`regexpert` breaks pattern matching into a plain-English pipeline:

1. **Build** your pattern using descriptive `xp_build_*` functions.
2. **Operate** on the pattern to specify repetitions or boundaries.
3. **Act** on it using `xp_action_find()` to return matches.

These steps are connected with the pipe (`%>%` or `|>`), making your code read like a sentence:
```{r eval = FALSE}
xp_build_digits() %>%
  xp_op_repeat(1, Inf) %>%
  xp_action_find("Call us at 800-555-1234")
#> [1] "800" "555" "1234"
```

No regex required. You described what you wanted, and `regexpert` handled the escaping and logic.

## The Builders

### `xp_build_digits()`

Matches numeric characters (0–9). In `regexpert`, builders match a single character by default. To match groups of digits, follow them with `xp_op_repeat()`.
```{r eval = FALSE}
xp_build_digits() %>%
  xp_op_repeat(1, Inf) %>%
  xp_action_find("Order #88412 placed on 2024-01-15")
#> [1] "88412" "2024" "01" "15"
```

### `xp_build_letters()`

Matches alphabetic characters (a–z, A–Z). You can specify the case if needed.
```{r eval = FALSE}
xp_build_letters(case = "both") %>%
  xp_op_repeat(1, Inf) %>%
  xp_action_find("Order #88412 placed")
#> [1] "Order" "placed"
```

### `xp_build_whitespace()`

Matches spaces, tabs, and newlines.
```{r eval = FALSE}
xp_build_whitespace() %>%
  xp_op_repeat(1, Inf) %>%
  xp_action_find("hello   world")
#> [1] "   "
```

### `xp_build_literal()`

Sometimes you need to match specific symbols like `$` or `.`. Usually, these are "dangerous" in regex, but `xp_build_literal()` automatically escapes them for you.
```{r eval = FALSE}
xp_build_literal("$") %>%
  xp_build_digits() %>%
  xp_op_repeat(1, Inf) %>%
  xp_action_find("The price is $100")
#> [1] "$100"
```

## Inspecting Your Work

If you ever want to see the actual regex string being generated under the hood, use `xp_action_view()`.
```{r eval = FALSE}
xp_build_digits() %>%
  xp_op_repeat(3) %>%
  xp_build_literal("-") %>%
  xp_action_view()
#> Current Regex Pattern:
#> (?:\d){3}\-
```