---
title: "Getting Started with regexpert"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with regexpert}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(regexpert)
library(magrittr)
```

## The Problem with Regular Expressions

If you've ever needed to extract a phone number or pull out specific text in R, you've probably run into something like this:
```{r eval = FALSE}
stringr::str_extract_all("Call us at 800-555-1234", "\\d+")
```

That `"\\d+"` is a regular expression. It works — but what does it *mean*? For anyone who doesn't have regex memorized, this is a wall. You have to stop what you're doing, look up the syntax, and hope you get the escaping right.

This is the problem `regexpert` solves.

## The Build → Operate → Act Workflow

`regexpert` breaks pattern matching into a plain-English pipeline:

1. **Build** your pattern using descriptive `xp_build_*` functions.
2. **Operate** on the pattern to specify repetitions or boundaries.
3. **Act** on it using `xp_action_find()` to return matches.

These steps are connected with the pipe (`%>%` or `|>`), making your code read like a sentence:
```{r eval = FALSE}
xp_build_digits() %>%
  xp_op_repeat(1, Inf) %>%
  xp_action_find("Call us at 800-555-1234")
#> [1] "800" "555" "1234"
```

No regex required. You described what you wanted, and `regexpert` handled the escaping and logic.

## The Builders

### `xp_build_digits()`

Matches numeric characters (0–9). In `regexpert`, builders match a single character by default. To match groups of digits, follow them with `xp_op_repeat()`.
```{r eval = FALSE}
xp_build_digits() %>%
  xp_op_repeat(1, Inf) %>%
  xp_action_find("Order #88412 placed on 2024-01-15")
#> [1] "88412" "2024" "01" "15"
```

### `xp_build_letters()`

Matches alphabetic characters (a–z, A–Z). You can specify the case if needed.
```{r eval = FALSE}
xp_build_letters(case = "both") %>%
  xp_op_repeat(1, Inf) %>%
  xp_action_find("Order #88412 placed")
#> [1] "Order" "placed"
```

### `xp_build_whitespace()`

Matches spaces, tabs, and newlines.
```{r eval = FALSE}
xp_build_whitespace() %>%
  xp_op_repeat(1, Inf) %>%
  xp_action_find("hello   world")
#> [1] "   "
```

### `xp_build_literal()`

Sometimes you need to match specific symbols like `$` or `.`. These are "dangerous" in regex, but `xp_build_literal()` automatically escapes them for you.
```{r eval = FALSE}
xp_build_literal("$") %>%
  xp_build_digits() %>%
  xp_op_repeat(1, Inf) %>%
  xp_action_find("The price is $100")
#> [1] "$100"
```

## Chaining Builders Together

The real power of `regexpert` comes from chaining multiple builders into a single pipeline. Each builder adds to the pattern left to right, letting you describe complex matches in plain English steps.

For example, say you want to extract prices from a messy string — a `$` sign followed by one or more digits, a dot, then exactly two decimal digits:
```{r eval = FALSE}
xp_build_literal("$") %>%
  xp_build_digits() %>%
  xp_op_repeat(1, Inf) %>%
  xp_build_literal(".") %>%
  xp_build_digits() %>%
  xp_op_repeat(2, 2) %>%
  xp_action_find("Items: $4.99, $12.50, and $120.00")
#> [1] "$4.99"   "$12.50"  "$120.00"
```

Or say you want to pull out hyphenated product codes — three letters, a dash, then four digits:
```{r eval = FALSE}
xp_build_letters(case = "upper") %>%
  xp_op_repeat(3, 3) %>%
  xp_build_literal("-") %>%
  xp_build_digits() %>%
  xp_op_repeat(4, 4) %>%
  xp_action_find("Products: ABC-1234, XYZ-5678, and DEF-9999")
#> [1] "ABC-1234" "XYZ-5678" "DEF-9999"
```

You can always use `xp_action_view()` to peek at the regex being built at any point in the pipeline — useful for debugging or learning:
```{r eval = FALSE}
xp_build_letters(case = "upper") %>%
  xp_op_repeat(3, 3) %>%
  xp_build_literal("-") %>%
  xp_build_digits() %>%
  xp_op_repeat(4, 4) %>%
  xp_action_view()
#> Current Regex Pattern:
#> [A-Z]{3}\-\d{4}
```

## Standard Patterns

For common real-world data types, `regexpert` includes a library of pre-built patterns via `xp_build_standard()`. No need to remember complex regex — just name what you want.
```{r eval = FALSE}
# Extract email addresses
xp_build_standard("email") %>%
  xp_action_find("Contact us at support@regexpert.com or sales@example.org")
#> [1] "support@regexpert.com" "sales@example.org"

# Extract phone numbers
xp_build_standard("phone") %>%
  xp_action_find("Call us at (208) 555-1234 or 208.555.5678")
#> [1] "(208) 555-1234" "208.555.5678"

# Extract ZIP codes
xp_build_standard("zip") %>%
  xp_action_find("My zip code is 83440")
#> [1] "83440"

# Extract ISO dates
xp_build_standard("date_iso") %>%
  xp_action_find("The event is on 2026-03-17")
#> [1] "2026-03-17"
```

The full list of available types includes: `email`, `phone`, `zip`, `date_iso`, `date_us`, `credit_card`, `ipv4`, `hex_color`, `time_24`, `ssn`, and `mac_address`.

Need a pattern that isn't in the library? Use `xp_register()` to add your own custom patterns to the session library and access them the same way via `xp_build_standard()`.

## Inspecting Your Work

If you ever want to see the actual regex string being generated under the hood, use `xp_action_view()`.
```{r eval = FALSE}
xp_build_digits() %>%
  xp_op_repeat(3) %>%
  xp_build_literal("-") %>%
  xp_action_view()
#> Current Regex Pattern:
#> (?:\d){3}\-
```

## Try It Yourself

Now that you've seen how `regexpert` works, here are a few challenges to test your understanding. Each one can be solved by chaining builders and operators together.

**Challenge 1 — Extract the time**

Given the string `"Meeting at 09:30 and follow-up at 14:45"`, write a pipeline that extracts both times in `HH:MM` format.

*Hint: a time is two digits, a literal colon, then two more digits.*

**Challenge 2 — Pull out the serial numbers**

Given `"Units: SN-00123, SN-00456, SN-00789"`, write a pipeline that extracts each serial number in `SN-XXXXX` format.

*Hint: think about what `xp_build_literal()` and `xp_op_repeat()` can do together.*

**Challenge 3 — Find the hex color codes**

Given `"Theme colors: #ff5733, #c70039, and #900c3f"`, try extracting the hex codes using `xp_build_standard()` first, then try building the same pattern manually from scratch using `xp_build_literal()` and `xp_build_alnum()`.

*Hint: a hex color is a `#` followed by exactly 6 alphanumeric characters.*

---

Once you're comfortable with the basics, explore the full function reference at `?xp_build_standard` and `?xp_op_repeat` to see everything `regexpert` can do.